Indicators for Ecosystem Condition in Norway

### 2023-06-08 alien_species.Rmd
Render book with trophic level biomass

---
title: "Alien species"
output:
  html_document
---
# Alien species {#alien-species}

@@

```{r setup, include=FALSE}
library(knitr)
library(tidyverse)
library(sf)
library(tmap)
library(ggpubr)
library(DT)
library(units)
library(eaTools)
library(tibble)
library(stars)
knitr::opts_chunk$set(echo = TRUE)

>>>>

# Alien species {#alien-species}

@@

```{r setup, include=FALSE}
library(knitr)
library(stringr, lib.loc = "/usr/lib/R/site-library")
library(tidyverse, lib.loc = "/usr/local/lib/R/site-library")
#library(tidyverse)
library(sf)
library(tmap)
library(ggpubr)
library(DT)
library(units)
library(eaTools)
library(tibble)
library(stars)
knitr::opts_chunk$set(echo = TRUE)

### 2023-06-08 climate_indicators_explained.Rmd
Render book with trophic level biomass

```{r setup, include=FALSE}
library(knitr)
knitr::opts_chunk$set(echo = TRUE)
#library(tidyverse) # cant update to v2 on server
library(dplyr)
library(ggplot2)
library(stars)
library(sf)
library(tmap)
library(lubridate)
library(ggpubr)
library(microbenchmark)
library(tictoc)
library(ncmeta)
library(parallel)
library(readr)
library(exactextractr)
library(tidyr)
library(data.table)
library(terra)
dir <- substr(getwd(), 1,2)

@@

This chapters describes the development of a workflow for generating or preparing indicators based on interpolated climate data from [SeNorge](https://senorge.no/). I describe the process using dummy data and also discuss alternate approaches and dead ends that occured in the development process. For a succinct workflow example for a spesific indikcator, see [Median summer temperature](#median-summer-temp).

@@

```{r, fig.cap="Five accounting areas (regions) in Norway."}

tm_shape(reg) +
  tm_polygons(col="region")

@@

### 1. Collate variable data series

@@

#### Filter data by dates

@@

#### Aggregate across time within a year

@@

These files are about 500 MB, but the tiff file reads in a lot quicker then the rds file, so from now on I'll just export as tiff. *(This could perhaps have been solved with a `gc()`,. Later I save as RData with no problems).*

```{r}
summer_median_temp <- read_stars("/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/climate_indicators/aggregated_climate_time_series/real_temp_summer.tiff",
                                 proxy=F)
# This is too slow:
# readRDS("/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/climate_indicators/dummy_data/summer_median_temp.rds")
# but it could be that RData works better 
```

@@

### 2. Calculate reference values

@@

```{r, eval=F}
# write_stars can only export a single attribute 
#write_stars(summer_median_temp_ref, "/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/climate_indicators/aggregated_climate_time_series/summer_median_temp_ref.tiff")

saveRDS(summer_median_temp_ref, "/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/climate_indicators/aggregated_climate_time_series/summer_median_temp_ref.RData")
```

```{r, eval=T, include=F}
summer_median_temp_ref <- readRDS("/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/climate_indicators/aggregated_climate_time_series/summer_median_temp_ref.RData")
```

@@

```{r, eval=T}
summer_median_temp_ref_long <- summer_median_temp_ref %>% split("median_sd")
```

@@

```{r}
y_var <- summer_median_temp %>%
  split(3) %>%
  c(summer_median_temp_ref_long)

#### Normalise variable

@@

@@

```{r, include=FALSE}
y_var_norm <- readRDS("/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/climate_indicators/aggregated_climate_time_series/temp_median_summer_normalised.RData")
```

@@

#### Plot time series as line graphs

>>>>

```{r setup, include=FALSE}
library(knitr)
knitr::opts_chunk$set(echo = TRUE)
library(stringr, lib.loc = "/usr/lib/R/site-library")
library(tidyverse, lib.loc = "/usr/local/lib/R/site-library")
#library(tidyverse) 
library(stars)
library(sf)
library(tmap)
library(lubridate)
library(ggpubr)
library(microbenchmark)
library(tictoc)
library(ncmeta)
library(parallel)
library(exactextractr)
library(data.table)
library(terra)
dir <- substr(getwd(), 1,2)
pData <- ifelse(dir == "C:", 
  "P:/41201785_okologisk_tilstand_2022_2023/data/",
  "/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/")
```

@@

This chapters describes the development of a workflow for generating or preparing indicators based on interpolated climate data from [SeNorge](https://senorge.no/). I describe the process using dummy data and also discuss alternate approaches and dead ends that occured in the development process. For a succinct workflow example for a specific indicator, see [Median summer temperature](#median-summer-temp).

@@

```{r, fig.cap="Five accounting areas (regions) in Norway."}
tm_shape(reg) +
  tm_polygons(col="region")
```

@@

### Collate variable data series

@@

### Filter data by dates

@@

### Aggregate across time within a year

@@

These files are about 500 MB, but the tiff file reads in a lot quicker then the rds file, so from now on I'll just export as tiff. *(This could perhaps have been solved with a `gc()`. Later I tried saving as RData with no speed issues problems, but then I had a problem with RData files seemingly becoming corrupted over time - stars objects saved as RData files would not bind together using `c.stars()` with other matching stars object and I think this could be related to a time stamp in the RData format. So, stick to tiff).*

```{r}
summer_median_temp <- read_stars(paste0(pData, "climate_indicators/aggregated_climate_time_series/real_temp_summer.tiff"),
                                 proxy=F)
```

@@

### Calculate reference values

@@

```{r, eval=F}
# Note that write_stars can only export a single attribute, but we have put the 'attributes' as dimensions, and these will be read in as 'band' when importing tiff.
write_stars(summer_median_temp_ref, "/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/climate_indicators/aggregated_climate_time_series/summer_median_temp_ref.tiff")

# I initially saved as RData, like this, but ran into problems later when trying to bind stars objects using c.stars(). It appears RData becomes corrutpt in a sense, over time, perhaps due to some time stamp. If I recreated the object in the global environment and saved as RDatam I could read it in and work with it, but the next day it would not work again.
#saveRDS(summer_median_temp_ref, "/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/climate_indicators/aggregated_climate_time_series/summer_median_temp_ref.RData")
```

```{r, eval=T, include=F}
summer_median_temp_ref <- read_stars(paste0(pData, "climate_indicators/aggregated_climate_time_series/summer_median_temp_ref.tiff"),
                                     proxy=F)
```

@@

```{r, eval=T}
summer_median_temp_ref_long <- summer_median_temp_ref %>% split("band")
```

@@

```{r}
y_var <- summer_median_temp %>%
  split(3) %>%
  c(summer_median_temp_ref_long)
```

### Normalise variable

@@

Writing to file, this time as RData (could use rds) because tiff don't allow multiple attributes, and because we don't need to use `c.stars()` in this object.

@@

```{r, include=FALSE}
y_var_norm <- readRDS(paste0(pData, "climate_indicators/aggregated_climate_time_series/temp_median_summer_normalised.RData"))
```

@@

### Plot time series as line graphs

### 2023-06-08 functional_plant_indicators_wetland.Rmd
Render book with trophic level biomass

<!-- Load all you dependencies here -->
```{r setup, include=FALSE}
library(sf)
library(plyr)
library(tidyverse)
library(RColorBrewer)
library("gridExtra") 
library(ggridges)
library(tmap)
library(tmaptools)
library(raster)
library(stars)
library(knitr)
knitr::opts_chunk$set(echo = TRUE)
```

>>>>

<!-- Load all you dependencies here -->
```{r setup, include=FALSE}
library(sf)
library(plyr)
library(stringr, lib.loc = "/usr/lib/R/site-library")
library(tidyverse, lib.loc = "/usr/local/lib/R/site-library")
#library(tidyverse)
library(RColorBrewer)
library("gridExtra") 
library(ggridges)
library(tmap)
library(tmaptools)
library(raster)
library(stars)
library(knitr)
knitr::opts_chunk$set(echo = TRUE)
```

### 2023-06-08 homogeneous_impact_areas.Rmd
Render book with trophic level biomass

---
title: "-"
output:
  html_document
editor_options: 
  markdown: 
    wrap: sentence
---

@@

```{r setup, include=FALSE}
library(knitr)
knitr::opts_chunk$set(echo = TRUE)
library(sf)
library(tmap)
library(stars);packageVersion("stars")
library(units)
library(ggpubr)
library(DT)
library(tidyverse)
library(units)
```

>>>>

@@

```{r setup, include=FALSE}
library(knitr)
knitr::opts_chunk$set(echo = TRUE)
library(sf)
library(tmap)
library(stars);packageVersion("stars")
library(units)
library(ggpubr)
library(DT)
library(stringr, lib.loc = "/usr/lib/R/site-library")
library(tidyverse, lib.loc = "/usr/local/lib/R/site-library")
#library(tidyverse)
library(ggplot2)
library(dplyr)
library(units)
```

### 2023-06-08 index.Rmd
Render book with trophic level biomass

```{r eaTools-example, fig.cap="A normalised condition variable is called an indicator. This indicator had been transformed with a sigmoid function."}
#devtools::install_github("NINAnor/eaTools")
library(eaTools)

@@

Kolstad, A.L., Topper, J. Grainger, M.J., Sandvik, H. Indicators for Ecosystem Condition in Norway. https://ninanor.github.io/ecosystemCondition/index.html.doi [TBA]

## License

>>>>

```{r workaround-lib-issue, echo=FALSE, warning=FALSE, message=FALSE}
library(stringr, lib.loc = "/usr/lib/R/site-library")
library(tidyverse, lib.loc = "/usr/local/lib/R/site-library")
```

```{r eaTools-example, fig.cap="A normalised condition variable is called an indicator. This indicator had been transformed with a sigmoid function."}
#devtools::install_github("NINAnor/eaTools")

library(eaTools)

@@

Kolstad, A.L., Topper, J. Grainger, M.J., Sandvik, H. (2023) Indicators for Ecosystem Condition in Norway. https://doi.org/10.5281/zenodo.7854835

## License {.unnumbered}

### 2023-06-08 medianSummerTemperature.Rmd
Render book with trophic level biomass

# Median Summer Temperature

@@

```{r setup, include=FALSE}
library(knitr)
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse) # cant update to v2 on server
library(stars)
library(sf)
library(tmap)
library(lubridate)
library(ggpubr)
library(microbenchmark)
library(tictoc)
library(ncmeta)
library(parallel)
library(exactextractr)
library(data.table)
library(terra)
dir <- substr(getwd(), 1,2)
```

@@

This chapters describes the workflow for an indicator describing the median summer temperature. For a more comprehensive documentaion on the development of the workflow itself, se [here](#climate-indicators-explained). The data comes from interpolated climate surfaces from [SeNorge](https://senorge.no/) which contain one 1x1km raster for each day since 1957 to present. The reference levels are extracted from the time period 1961-1990 (a temporal reference condition).

@@

```{r, eval=F}
# a backup caching in case tiff causes problems down the line
saveRDS(real_temp_summer, "/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/climate_indicators/aggregated_climate_time_series/real_temp_summer.rds")

```{r, include=F}
real_temp_summer <- read_stars("/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/climate_indicators/aggregated_climate_time_series/real_temp_summer.tiff",
                                 proxy=F)
```

@@

<!-- RUN AGAIN -->
```{r, eval=FALSE}
system.time({
cl <- makeCluster(10L)
summer_median_temp_ref <- summer_median_temp_ref %>%
  st_apply(c("x", "y"), 
           FUN =  median_sd,
           CLUSTER = cl)
stopCluster(cl)
})
```

@@

```{r, eval=F}
saveRDS(summer_median_temp_ref, "/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/climate_indicators/aggregated_climate_time_series/summer_median_temp_ref.RData")
```

```{r, eval=T, include=F}
summer_median_temp_ref <- readRDS("/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/climate_indicators/aggregated_climate_time_series/summer_median_temp_ref.RData")
```

@@

```{r, eval=T}
summer_median_temp_ref_long <- summer_median_temp_ref %>% 
  split("median_sd") %>%
  setNames(c("reference_upper", "sd"))
```

@@

```{r, eval=F}
gc()
saveRDS(y_var_norm, "/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/climate_indicators/aggregated_climate_time_series/summer_median_temp_normalised.RData")
```

```{r, include=FALSE}
y_var_norm <- readRDS("/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/climate_indicators/aggregated_climate_time_series/temp_median_summer_normalised.RData")
```

>>>>

# Median Summer Temperature {#median-summer-temp}

@@

```{r setup, include=FALSE}
library(knitr)
knitr::opts_chunk$set(echo = TRUE)
library(stringr, lib.loc = "/usr/lib/R/site-library")
library(tidyverse, lib.loc = "/usr/local/lib/R/site-library")
#library(tidyverse)
library(stars)
library(sf)
library(tmap)
library(lubridate)
library(ggpubr)
library(microbenchmark)
library(tictoc)
library(ncmeta)
library(parallel)
library(exactextractr)
library(data.table)
library(terra)
dir <- substr(getwd(), 1,2)
pData <- ifelse(dir == "C:", 
  "P:/41201785_okologisk_tilstand_2022_2023/data/",
  "/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/")
```

@@

This chapters describes the workflow for an indicator describing the median summer temperature. For a more comprehensive documentation on the development of the workflow itself, see [here](#climate-indicators-explained). The data comes from interpolated climate surfaces from [SeNorge](https://senorge.no/) which contain one 1x1km raster for each day since 1957 to present. The reference levels are extracted from the time period 1961-1990 (a temporal reference condition).

@@

```{r, include=F}
real_temp_summer <- read_stars(paste0(pData, "climate_indicators/aggregated_climate_time_series/real_temp_summer.tiff"),
                                 proxy=F)
```

@@

```{r, eval=FALSE}
system.time({
cl <- makeCluster(10L)
summer_median_temp_ref <- summer_median_temp_ref %>%
  st_apply(c("x", "y"), 
           FUN =  median_sd,
           CLUSTER = cl)
stopCluster(cl)
})
```

@@

```{r, eval=F}
write_stars(summer_median_temp_ref, "/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/climate_indicators/aggregated_climate_time_series/summer_median_temp_ref.tiff")
```

```{r, eval=T, include=F}
summer_median_temp_ref <- read_stars(paste0(pData, "climate_indicators/aggregated_climate_time_series/summer_median_temp_ref.tiff"),
                                     proxy=F)
```

@@

```{r, eval=T}
summer_median_temp_ref_long <- summer_median_temp_ref %>% 
  split("band") %>%
  setNames(c("reference_upper", "sd"))
```

@@

```{r, eval=F}
gc()
saveRDS(y_var_norm, "/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/climate_indicators/aggregated_climate_time_series/summer_median_temp_normalised.RData")

# Tiff dont allow for multiple attributes:
#write_stars(y_var_norm, "/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/climate_indicators/aggregated_climate_time_series/summer_median_temp_normalised.tiff")
```

```{r, include=FALSE}
y_var_norm <- readRDS(paste0(pData, "climate_indicators/aggregated_climate_time_series/temp_median_summer_normalised.RData"))
```
